Some examples in extended logic counting classes

### 2017-04-07 LICS/extentions.tex
minor corrections

There exist complexity classes that do not fit in our framework because either the output of a function in not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max are used).
To remedy this problem, we show here how $\qso$ can be easily extended  to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} the author does not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} the authors define the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
%operators of first- and second-order. 
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
%the formulae $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. 
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

There exist complexity classes that do not fit in our framework because either the output of a function is not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max are used).
To remedy this problem, we show here how $\qso$ can be easily extended to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Just as we did for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust $\qso$ is when capturing different complexity classes.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems are.
Given that in~\cite{krentel1988complexity} the author does not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} the authors define the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguish between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
%operators of first- and second-order. 
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
%the formulae $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. 
As one could expect, $\maxqso$ and $\minqso$ are the logics needed to capture $\maxp$ and $\minp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

### 2017-04-07 LICS/framework.tex
minor corrections

Let $\R$ be a signature, $\A$ an $\R$-structure with domain $A$, $v$ a first-order assignment for $\A$ and $V$ a second-order assignment for $\A$. Then the \emph{evaluation} of a $\qso$-formula $\alpha$ over $(\A, v, V)$ is defined as a function $\sem{\alpha}$ that on input $(\A, v, V)$ returns a number in $\bbN$. Formally, the function $\sem{\alpha}$ is recursively defined in Table~\ref{tab-semantics}.
\begin{table}
	\addtolength{\jot}{0.5em}
	\begin{align*}
	\sem{\varphi}(\A, v, V) & = 
	\begin{cases}
	1 & \mbox{if } (\A, v, V) \models \varphi \\
	0 & \mbox{otherwise}
	\end{cases}\\
	\sem{s}(\A, v, V) & = s \\
	\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
	\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
	\sem{\sa{x} \alpha}(\A, v, V) & = \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
	\sem{\pa{x} \alpha}(\A, v, V) & = \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
	\sem{\sa{X} \alpha}(\A, v, V) & = \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
	\sem{\pa{X} \alpha}(\A, v, V) & = \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
	\end{align*}
	\caption{The semantics of QSO formulae.}
	\label{tab-semantics}
	\vspace*{-21pt}
\end{table}
%$$
%\renewcommand{\arraystretch}{1.7}
%\begin{array}{rcl} 
%\sem{\varphi}(\A, v, V) & = & 
%\begin{cases}
%1 & \mbox{if } (\A, v, V) \models \varphi \\
%0 & \mbox{otherwise}
%\end{cases}\\
%\sem{s}(\A, v, V) & = & s \\
%\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
%\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
%\sem{\sa{x} \alpha}(\A, v, V) & = & \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
%\sem{\pa{x} \alpha}(\A, v, V) & = & \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
%\sem{\sa{X} \alpha}(\A, v, V) & = & \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
%\sem{\pa{X} \alpha}(\A, v, V) & = & \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
%\end{array}
%$$
A $\qso$-formula $\alpha$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\alpha$ is under the scope of a usual quantifier or a quantitative quantifier. It is important to notice that if $\alpha$ is a $\qso$-sentence over a signature $\R$, then for every $\R$-structure $\A$, first-order assignments $v_1$, $v_2$ for $\A$ and second-order assignments $V_1$, $V_2$ for $\A$, it holds that $\sem{\alpha}(\A, v_1, V_1) = \sem{\alpha}(\A, v_2, V_2)$.
Thus, in such a case we use the term $\sem{\alpha}(\A)$ to denote $\sem{\alpha}(\A, v, V)$, for some arbitrary first-order assignment $v$ for $\A$ and some arbitrary second-order assignment $V$ for $\A$. 
\begin{example}\label{ex:cliques}
Let $\bG = \{E(\cdot,\cdot),<\}$ be the vocabulary for graphs and $\fG$ be an ordered $\bG$-structure encoding a non-directed graph. 
Suppose that we want to count the number of triangles in $\fG$. Then this can be defined as follows:
\begin{multline*}
\alpha_1 \ := \ \sa{x} \sa{y} \sa{z} ( E(x,y) \, \wedge \, E(y,z) \, \wedge \, E(z,x) \, \wedge \\
x < y \, \wedge \, y < z )
\end{multline*}
We encode a triangle in $\alpha_1$ as an increasing sequence of nodes $\{x, y, z\}$, in order to count each triangle once. Then the Boolean subformula  $E(x,y) \wedge E(y,z) \wedge E(z,x) \wedge
x < y \wedge y < z$ is checking the triangle property, by returning $1$ if $\{x, y, z\}$ forms a triangle in $\fG$ and $0$ otherwise.
Finally, the sum quantifiers in $\alpha_1$ aggregates all the values, counting the number of triangles in $\fG$.

@@

Finally, the work in~\cite{SalujaST95} and \cite{ComptonG96} is of particular interest for our research. 
In~\cite{SalujaST95}, it was proposed to define a function over a structure by using free variables in an SO-formula; in particular, the function is defined by the number of instantiations of the free variables that are satisfied by the structure.
%and then counting how many extensions of an structure satisfy the formula (i.e. by instantiating the free variables). 
Formally, Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for a fragment $\LL$ of $\fo$. For a formula $\varphi(\bar{x},\bar{X})$ over $\R$, the function $f_{\varphi(\bar x, \bar X)}$ is defined as
$
f_{\varphi(\bar x, \bar X)}(\A) = \vert \{(\bar{a},\bar{A}) \mid \A\models\varphi(\bar{a},\bar{A})\}\vert,
$
for every $\A\in\ostr[\R]$. Then a function $g:\ostr[\R]\to\nat$ is in $\#\LL$ if there exists a formula $\varphi(\bar{x},\bar{X})$ in $\LL$ such that $g = f_{\varphi(\bar x, \bar X)}$.
%\martin{La notacion $f_{\psi(x,X)}$ la inventamos}
In~\cite{SalujaST95}, they were proved several results about capturing counting complexity classes which are relevant for our work. We discuss and use these results in Sections~\ref{sec:complexity} and~\ref{sec:syntactic}.
Notice that for every formula $\varphi(\bar{x},\bar{X})$, it holds that $f_{\varphi(\bar{x},\bar{X})}$ is the same function as $\sem{\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{x},\bar{X})}$, that is, the approach in \cite{SalujaST95} can be seen as a syntactical restriction of our approach based on $\qso$. 
Thus, the advantage of our approach relies on the flexibility to define functions by alternating sum with product operators and, moreover, by introducing new quantitative operators (see Section~\ref{sec:beyond}).
Furthermore, we show in the following section how to capture some fundamental classes that cannot be captured by following the approach in~\cite{SalujaST95}.

>>>>

Let $\R$ be a signature, $\A$ an $\R$-structure with domain $A$, $v$ a first-order assignment for $\A$ and $V$ a second-order assignment for $\A$. Then the \emph{evaluation} of a $\qso$-formula $\alpha$ over $(\A, v, V)$ is defined as a function $\sem{\alpha}$ that on input $(\A, v, V)$ returns a number in $\bbN$. Formally, the function $\sem{\alpha}$ is recursively defined in Table~\ref{tab-semantics}.
\begin{table}
	\addtolength{\jot}{0.5em}
	\begin{align*}
	\sem{\varphi}(\A, v, V) & = 
	\begin{cases}
	1 & \mbox{if } (\A, v, V) \models \varphi \\
	0 & \mbox{otherwise}
	\end{cases}\\
	\sem{s}(\A, v, V) & = s \\
	\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
	\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
	\sem{\sa{x} \alpha}(\A, v, V) & = \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
	\sem{\pa{x} \alpha}(\A, v, V) & = \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
	\sem{\sa{X} \alpha}(\A, v, V) & = \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
	\sem{\pa{X} \alpha}(\A, v, V) & = \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
	\end{align*}
	\caption{The semantics of QSO formulae.}
	\label{tab-semantics}
	\vspace*{-21pt}
\end{table}
%$$
%\renewcommand{\arraystretch}{1.7}
%\begin{array}{rcl} 
%\sem{\varphi}(\A, v, V) & = & 
%\begin{cases}
%1 & \mbox{if } (\A, v, V) \models \varphi \\
%0 & \mbox{otherwise}
%\end{cases}\\
%\sem{s}(\A, v, V) & = & s \\
%\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
%\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
%\sem{\sa{x} \alpha}(\A, v, V) & = & \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
%\sem{\pa{x} \alpha}(\A, v, V) & = & \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
%\sem{\sa{X} \alpha}(\A, v, V) & = & \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
%\sem{\pa{X} \alpha}(\A, v, V) & = & \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
%\end{array}
%$$
A $\qso$-formula $\alpha$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\alpha$ is under the scope of a usual quantifier or a quantitative quantifier. It is important to notice that if $\alpha$ is a $\qso$-sentence over a signature $\R$, then for every $\R$-structure $\A$, first-order assignments $v_1$, $v_2$ for $\A$ and second-order assignments $V_1$, $V_2$ for $\A$, it holds that $\sem{\alpha}(\A, v_1, V_1) = \sem{\alpha}(\A, v_2, V_2)$.
Thus, in such a case we use the term $\sem{\alpha}(\A)$ to denote $\sem{\alpha}(\A, v, V)$, for some arbitrary first-order assignment $v$ for $\A$ and some arbitrary second-order assignment $V$ for $\A$. 
\begin{example}\label{ex:cliques}
Let $\bG = \{E(\cdot,\cdot),<\}$ be the vocabulary for graphs and $\fG$ be an ordered $\bG$-structure encoding an undirected graph. 
Suppose that we want to count the number of triangles in $\fG$. Then this can be defined as follows:
\begin{multline*}
\alpha_1 \ := \ \sa{x} \sa{y} \sa{z} ( E(x,y) \, \wedge \, E(y,z) \, \wedge \, E(z,x) \, \wedge \\
x < y \, \wedge \, y < z )
\end{multline*}
We encode a triangle in $\alpha_1$ as an increasing sequence of nodes $\{x, y, z\}$, in order to count each triangle once. Then the Boolean subformula  $E(x,y) \wedge E(y,z) \wedge E(z,x) \wedge
x < y \wedge y < z$ is checking the triangle property, by returning $1$ if $\{x, y, z\}$ forms a triangle in $\fG$ and $0$ otherwise.
Finally, the sum quantifiers in $\alpha_1$ aggregates all the values, counting the number of triangles in $\fG$.

@@

Finally, the work in~\cite{SalujaST95} and \cite{ComptonG96} is of particular interest for our research. 
In~\cite{SalujaST95}, it was proposed to define a function over a structure by using free variables in an SO-formula; in particular, the function is defined by the number of instantiations of the free variables that are satisfied by the structure.
%and then counting how many extensions of an structure satisfy the formula (i.e. by instantiating the free variables). 
Formally, Saluja et. al \cite{SalujaST95} define a family of counting classes $\#\LL$ for a fragment $\LL$ of $\fo$. For a formula $\varphi(\bar{x},\bar{X})$ over $\R$, the function $f_{\varphi(\bar x, \bar X)}$ is defined as
$
f_{\varphi(\bar x, \bar X)}(\A) = \vert \{(\bar{a},\bar{A}) \mid \A\models\varphi(\bar{a},\bar{A})\}\vert,
$
for every $\A\in\ostr[\R]$. Then a function $g:\ostr[\R]\to\nat$ is in $\#\LL$ if there exists a formula $\varphi(\bar{x},\bar{X})$ in $\LL$ such that $g = f_{\varphi(\bar x, \bar X)}$.
%\martin{La notacion $f_{\psi(x,X)}$ la inventamos}
In~\cite{SalujaST95}, several results were proved about capturing counting complexity classes which are relevant for our work. We discuss and use these results in Sections~\ref{sec:complexity} and~\ref{sec:syntactic}.
Notice that for every formula $\varphi(\bar{x},\bar{X})$, it holds that $f_{\varphi(\bar{x},\bar{X})}$ is the same function as $\sem{\sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{x},\bar{X})}$, that is, the approach in \cite{SalujaST95} can be seen as a syntactical restriction of our approach based on $\qso$. 
Thus, the advantage of our approach relies on the flexibility to define functions by alternating sum with product operators and, moreover, by introducing new quantitative operators (see Section~\ref{sec:beyond}).
Furthermore, we show in the following section how to capture some fundamental classes that cannot be captured by following the approach in~\cite{SalujaST95}.

### 2017-04-07 LICS/preliminaries.tex
minor corrections

From now on, assume given disjoint infinite sets $\fv$ and $\sv$ of first-order variables and second-order variables, respectively. Notice that every variable in $\sv$ has an associated arity, which is denoted by $\arity(X)$. Then given a  signature $\R$, the set of second-order logic formulae ($\so$-formulae) over $\R$ is given by the following grammar:
\begin{multline*}\ 
	\varphi \ := \ x = y \ \mid \ R(\bar u) \ \mid \ \top  \ \mid\  
	X(\bar v)  \ \mid\\ 
	\neg \varphi \ \mid\ 
	(\varphi \vee \varphi) \ \mid\ 
	\exists x \, \varphi \ \mid\ 
	\exists X \, \varphi
\end{multline*}
where $x,y \in \fv$, $R \in \R$, $\bar u$ is a tuple of (non-necessarily distinct) variables from $\fv$ whose length is $\arity(R)$, $\top$ is a reserved symbol to represent a tautology, $X \in \sv$, $\bar v$ is a tuple of (non-necessarily distinct) variables from $\fv$ whose length is $\arity(X)$, and $x \in \fv$.

>>>>

From now on, assume given disjoint infinite sets $\fv$ and $\sv$ of first-order variables and second-order variables, respectively. Notice that every variable in $\sv$ has an associated arity, which is denoted by $\arity(X)$. Then given a  signature $\R$, the set of second-order logic formulae ($\so$-formulae) over $\R$ is given by the following grammar:
\begin{multline*}\ 
	\varphi \ := \ x = y \ \mid \ R(\bar u) \ \mid \ \top  \ \mid\  
	X(\bar v)  \ \mid\\ 
	\neg \varphi \ \mid\ 
	(\varphi \vee \varphi) \ \mid\ 
	\exists x \, \varphi \ \mid\ 
	\exists X \, \varphi
\end{multline*}
where $x,y \in \fv$, $R \in \R$, $\bar u$ is a tuple of (not necessarily distinct) variables from $\fv$ whose length is $\arity(R)$, $\top$ is a reserved symbol to represent a tautology, $X \in \sv$, $\bar v$ is a tuple of (not necessarily distinct) variables from $\fv$ whose length is $\arity(X)$, and $x \in \fv$.
